---
title: "Task shapes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Task shapes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# add_task() returns the task visibly. The render function prints every other
# value, and no R6 object.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  render = function(x, ...) {
    if (inherits(x, "R6")) return(invisible(NULL))
    knitr::knit_print(x, ...)
  }
)
```

A task shape is a plan design that recurs across tasks. It fixes how the plans split the work, what the first and the last plan do, and how the task writes. This vignette and `vignette("more-task-shapes")` describe five shapes. Each shape runs against SQLite when the vignette builds. A chunk calls `stop()` if a shape does not keep its promise, so the build fails. `vignette("how-a-task-runs")` states the execution rules that every shape follows.

| Shape | Use it when | Vignette |
|---|---|---|
| [Per-unit rebuild](more-task-shapes.html#shape-per-unit-rebuild) | the rows of each unit depend only on its own data pull, and no reader needs the table during the run | `more-task-shapes` |
| [Staged import](#shape-staged-import) | an import replaces the weeks of a run, and the live table MUST change only when every week arrived | `task-shapes` |
| [Staging as a checkpoint](more-task-shapes.html#shape-staging-checkpoint) | an import is long, and a failed run MUST NOT lose the weeks that it already pulled | `more-task-shapes` |
| [Fit then fill](#shape-fit-then-fill) | a fit can fail for one unit, and the fallback needs the results of all units | `task-shapes` |
| [Export from a finished table](more-task-shapes.html#shape-export-finished-table) | the output is files, made from tables that an earlier task finished | `more-task-shapes` |

## Setup

CS9 reads its database configuration from environment variables. This chunk points the `config`
and `anon` access levels at two SQLite files under `tempdir()`.

```{r setup}
library(data.table)

db_dir <- file.path(tempdir(), "cs9-task-shapes")
unlink(db_dir, recursive = TRUE)
dir.create(db_dir)
Sys.setenv(
  CS9_AUTO = "0",
  CS9_PATH = db_dir,
  CS9_DBCONFIG_ACCESS = "config/anon",
  CS9_DBCONFIG_DRIVER = "SQLite",
  CS9_DBCONFIG_DB_CONFIG = file.path(db_dir, "config.sqlite"),
  CS9_DBCONFIG_DB_ANON = file.path(db_dir, "anon.sqlite")
)
cs9::reload_db_config()

ss <- cs9::SurveillanceSystem_v9$new(name = "shapes")

# The rows of a table as a data.frame, sorted by the columns in `by`.
read_rows <- function(table, by) {
  d <- as.data.frame(dplyr::collect(table$tbl()))
  d <- d[do.call(order, d[by]), , drop = FALSE]
  rownames(d) <- NULL
  d
}
```

## Staged import {#shape-staged-import}

The import task has one plan per week. Each argset copies its week from the source into a staging
table. The argset with `first_analysis = TRUE` empties the staging table first. The argset with
`last_analysis = TRUE` checks the staged rows. It then swaps them into the live table in one
transaction.

- **Use it when** a run replaces a set of weeks. A reader of the live table then sees either the old weeks or the new weeks, never a mix.
- **Do not use it when** a run is long enough that a failure near the end costs a night of pulls. The first plan empties staging, so the next run pulls every week again. Use [staging as a checkpoint](more-task-shapes.html#shape-staging-checkpoint).
- **Failure modes.** Only the last plan MAY check and swap, because the middle plans MAY run in parallel, in any order ([rule](how-a-task-runs.html#rule-first-last)). A partitioned staging table MUST hold a week in every declared partition before the week counts as staged ([rule](how-a-task-runs.html#rule-partitions)).

```{r shape1-define}
weeks <- c("2026-01", "2026-02", "2026-03", "2026-04")
import_fields <- c(isoyearweek = "TEXT", location_code = "TEXT", n = "INTEGER")
for (variant in c("staging", "live")) {
  ss$add_table(
    name_access = "anon",
    name_grouping = "import",
    name_variant = variant,
    field_types = import_fields,
    keys = c("isoyearweek", "location_code")
  )
}

source_counts <- data.table(
  isoyearweek = rep(weeks, each = 2),
  location_code = c("county_03", "county_11"),
  n = 1:8
)
fail_week <- NA_character_

import_data <- function(argset, tables) {
  list(counts = source_counts[isoyearweek == argset$isoyearweek])
}

import_action <- function(data, argset, tables) {
  if (argset$first_analysis) tables$staging$drop_all_rows()
  if (identical(argset$isoyearweek, fail_week)) {
    stop("the source failed for ", fail_week)
  }
  if (nrow(data$counts) > 0) tables$staging$insert_data(data$counts)
  if (argset$last_analysis) swap_into_live(tables, argset$weeks)
}

# Deletes every week of the run from the live table, not only the weeks that
# reached staging, and appends the staged rows. One transaction holds both.
swap_into_live <- function(tables, weeks) {
  staged <- dplyr::collect(tables$staging$tbl())
  stopifnot(all(staged$isoyearweek %in% weeks), !anyNA(staged$n))
  live <- tables$live
  con <- live$dbconnection$autoconnection
  run_weeks <- paste(DBI::dbQuoteString(con, weeks), collapse = ", ")
  DBI::dbWithTransaction(con, {
    DBI::dbExecute(con, paste0(
      "DELETE FROM ", live$table_name_fully_specified_text,
      " WHERE isoyearweek IN (", run_weeks, ")"
    ))
    DBI::dbAppendTable(
      con,
      live$table_name_short_for_mssql_fully_specified_for_postgres,
      staged
    )
  })
}

ss$add_task(
  name_grouping = "import",
  name_action = "weekly",
  for_each_plan = lapply(weeks, function(w) list(isoyearweek = w)),
  universal_argset = list(weeks = weeks),
  data_selector_fn_name = "import_data",
  action_fn_name = "import_action",
  tables = list(
    staging = ss$tables$anon_import_staging,
    live = ss$tables$anon_import_live
  )
)
```

The live table holds an earlier run of all 4 weeks. The source then withdraws week 2026-03, so
no row for that week reaches staging. After the run, the live table MUST hold exactly the staged
rows, and no row for 2026-03.

```{r shape1-run}
ss$tables$anon_import_live$insert_data(
  data.table(isoyearweek = weeks, location_code = "county_03", n = 99L)
)
source_counts <- source_counts[isoyearweek != "2026-03"]

ss$run_task("import_weekly")
live <- read_rows(ss$tables$anon_import_live, c("isoyearweek", "location_code"))
staged <- read_rows(ss$tables$anon_import_staging, c("isoyearweek", "location_code"))
stopifnot(identical(live, staged), !"2026-03" %in% live$isoyearweek)
live[names(import_fields)]
```

The second run reads new counts, and week 2026-02 fails. The task MUST stop with an error, and
the live table MUST keep the rows of the first run.

```{r shape1-fail}
source_counts[, n := n + 100L]
fail_week <- "2026-02"
run <- tryCatch(ss$run_task("import_weekly"), error = function(e) e)
stopifnot(
  inherits(run, "error"),
  identical(
    read_rows(ss$tables$anon_import_live, c("isoyearweek", "location_code")),
    live
  )
)
conditionMessage(run)
cat("STAGED IMPORT CHECKS PASSED\n")
```

## Fit then fill {#shape-fit-then-fill}

The fit task fits a threshold for each stratum in its own argset. It writes `NA` for a stratum
with fewer than 4 weeks of history. The fill task has one plan. It fills each `NA` row from the
national row, records the `source` of each threshold, and publishes the whole result with
`csdb::DBTable_v9$replace_all_rows()`.

- **Use it when** a fit can fail for one unit, and its fallback needs the result of another unit, here the national row.
- **Do not use it when** every unit has a valid result of its own. A [per-unit rebuild](more-task-shapes.html#shape-per-unit-rebuild) is then enough.
- **Failure modes.** A middle plan of the fit task MUST NOT fill, because the national row MAY not exist yet ([rule](how-a-task-runs.html#rule-first-last)). The fill reads the whole fitted table in one pull, so the fitted table MUST stay small. The published table changes in one transaction, so a failed fill leaves the previous thresholds.

```{r shape2-define}
ss$add_table(
  name_access = "anon",
  name_grouping = "threshold",
  name_variant = "fitted",
  field_types = c(location_code = "TEXT", threshold = "DOUBLE"),
  keys = "location_code"
)
ss$add_table(
  name_access = "anon",
  name_grouping = "threshold",
  name_variant = "published",
  field_types = c(location_code = "TEXT", threshold = "DOUBLE", source = "TEXT"),
  keys = "location_code"
)

history <- data.table(
  location_code = rep(
    c("norge", "county_03", "county_11", "county_46"),
    times = c(8, 8, 8, 2)
  ),
  n = c(
    rep(c(40L, 44L, 38L, 46L), 2),
    rep(c(10L, 12L, 9L, 14L), 2),
    rep(c(20L, 25L, 18L, 22L), 2),
    c(3L, 4L)
  )
)
break_publish <- FALSE

fit_data <- function(argset, tables) {
  list(history = history[location_code == argset$location_code])
}

fit_action <- function(data, argset, tables) {
  n <- data$history$n
  threshold <- if (length(n) >= 4) mean(n) + 2 * sd(n) else NA_real_
  tables$fitted$upsert_data(
    data.table(location_code = argset$location_code, threshold = threshold)
  )
}

fill_data <- function(argset, tables) {
  list(fitted = as.data.table(dplyr::collect(tables$fitted$tbl())))
}

fill_action <- function(data, argset, tables) {
  d <- copy(data$fitted)
  national <- d[location_code == "norge", threshold]
  d[, source := fifelse(is.na(threshold), "national", "fitted")]
  d[is.na(threshold), threshold := national]
  # A duplicate key makes the publish fail on purpose.
  if (break_publish) d <- rbind(d, d[1])
  tables$published$replace_all_rows(d)
}

ss$add_task(
  name_grouping = "threshold",
  name_action = "fit",
  for_each_plan = lapply(unique(history$location_code), function(x) {
    list(location_code = x)
  }),
  data_selector_fn_name = "fit_data",
  action_fn_name = "fit_action",
  tables = list(fitted = ss$tables$anon_threshold_fitted)
)
ss$add_task(
  name_grouping = "threshold",
  name_action = "fill",
  for_each_plan = list(list()),
  data_selector_fn_name = "fill_data",
  action_fn_name = "fill_action",
  tables = list(
    fitted = ss$tables$anon_threshold_fitted,
    published = ss$tables$anon_threshold_published
  )
)
```

After both tasks run, every published row MUST have a threshold and a source. County 46 MUST
carry the national threshold.

```{r shape2-run}
ss$run_task("threshold_fit")
ss$run_task("threshold_fill")
published <- read_rows(ss$tables$anon_threshold_published, "location_code")
stopifnot(
  !anyNA(published$threshold),
  !anyNA(published$source),
  published$source[published$location_code == "county_46"] == "national"
)
published[c("location_code", "threshold", "source")]
```

The history then doubles, and the fit runs again. The publish of the fill task fails. The
published table MUST keep its previous contents.

```{r shape2-fail}
history[, n := 2L * n]
ss$run_task("threshold_fit")
break_publish <- TRUE
run <- tryCatch(ss$run_task("threshold_fill"), error = function(e) e)
stopifnot(
  inherits(run, "error"),
  identical(read_rows(ss$tables$anon_threshold_published, "location_code"), published)
)
conditionMessage(run)
cat("FIT THEN FILL CHECKS PASSED\n")
```

## Three rules {#three-rules}

**Publish in one transaction.** A task once emptied its MEM table in `first_analysis` and refilled
it argset by argset. Between those two points, and after any failed argset, readers saw an empty or
partial table. Fit then fill writes the published table in one call to `replace_all_rows()`.

**Replace the whole scope of the run.** A swap once deleted from the live table only the weeks that
each staging partition held. A week with no staged rows therefore kept its stale rows. The staged
import deletes every week of the run, so a withdrawn week leaves the live table.

**Record where each value came from.** A national threshold copied into a county row looks the same
as a threshold fitted for that county. Fit then fill writes `source` next to each threshold.

Both shapes need cs9 to run the argset with `last_analysis = TRUE` after every other argset of the
task. `tests/testthat/test-first-last-analysis.R` pins that order.
